Dialect: A Compact Line Notation Based on SMILES

### 2022-05-27 manuscript.tex
Re-works Goals to use molecular tree.

\section*{Goals, Design, and Tradeoffs}

@@

Dialect molecules inherit behavior from an underlying \textit{molecular graph}~\cite{balaban:1985}. A graph is a data structure comprised of a set of nodes and a set of pairwise relationships between them called edges. A molecular graph is a specialized graph onto which chemically-relevant metadata have been overlaid. The presence of this overlay allows nodes and edges to represent atoms and bonds, respectively. Molecular graphs are ubiquitous in cheminformatics, appearing in both serialization formats in-memory data structures found in software toolkits.

A Dialect molecule is a molecular graph having zero or more nodes and zero or more edges. The empty molecule, devoid of atoms and bonds, is therefore allowed. Dialect imposes an upper bound on the number of nodes of $2^{32} - 1$ and the same limit for the number of edges. However, practical limitations related to memory, storage, and CPU time will likely further constrain these limits in practice.

The edges of a Dialect molecular graph are defined as a set of unordered pairs of non-identical nodes. The practical consequences of this are threefold. First, \textit{loops} are not allowed. A loop is an edge joining a node to itself. Second, edges are \textit{undirected}. An undirected edge is equivalent to another edge whose terminals are transposed. With one exception that will be discussed in detail, an undirected edge between nodes A and B is therefore identical to an undirected edge between nodes B and A. Third, \textit{parallel edges} are disallowed. Edges are said to be parallel if they connect the same two nodes.

A molecular graph contains zero or more \textit{connected components}. A connected component is a graph in which any two vertices are connected to each other by at least one path. The maximum number of connected components is equal to the maximum node count, although practical factors may reduce this limit.

@@

Dialect uses several abstractions collectively known as the \textit{data model}. The data model defines metadata attributes associated with the components of a molecular graph. These attributes are themselves composed of \textit{primitives} (Table~\ref{table:primitives}), or irreducible data types.

\begin{table}
\begin{minipage}{\textwidth}
\renewcommand*\footnoterule{}
\caption{Data Model Primitives.}
\label{table:primitives}
\centering
\begin{tabular}{l l l}
    \hline
    Name & Description & Values\footnote{Ranges defined as \ttt{a..b} include \ttt{a} and exclude \ttt{b}.} \\
    \hline
    \ttt{Option<T>} & Optional Value & \ttt{None}, \ttt{T} \\
    \ttt{Index} & Unsigned integer & 0..2\textsuperscript{32} \\
    \ttt{Three} & Integer count & 1..4 \\
    \ttt{Dec} & Integer count &0..10 \\
    \ttt{SignedDec} & Integer count & -9..10 \\
    \ttt{Thou} & Integer count & 1..1000 \\
    \ttt{Element} & IUPAC-\-approved element symbol & \ttt{Ac}, \ttt{Ag}, \ttt{Al}, {\ldots}\ttt{Zn}\\
    \ttt{Configuration} & Configurational descriptor & \ttt{Right}, \ttt{Left} \\
    \ttt{Direction} & Conformational descriptor & \ttt{Up}, \ttt{Down} \\
    \ttt{Implicit} & Implicit hydrogen count flag & - \\
    \ttt{boolean} & Boolean & \ttt{true}, \ttt{false} \\
    \hline
\end{tabular}
\end{minipage}
\end{table}

@@

\begin{table}
\caption{Atomic State.}
\centering
\begin{tabular}{l L L l}
    \hline
    Attribute & Description & Type & Default \\
    \hline
    \ttt{index} & Unique identifier &\ttt{Index} & - \\
    \ttt{element} & Elemental symbol & \ttt{Option} \ttt{<Element>} & \ttt{None} \\
    \ttt{hydrogens} & Hydrogen count & \ttt{Implicit}, \ttt{Dec} & 0 \\
    \ttt{configuration} & Configuration & \ttt{Option} \ttt{<Configuration>} & \ttt{None} \\
    \ttt{charge} & Formal charge & \ttt{SignedDec} & 0 \\
    \ttt{isotope} & Mass number & \ttt{Option<Thou>} & \ttt{None} \\
    \ttt{selected} & Selection status & \ttt{boolean} & \ttt{false} \\
    \hline
\end{tabular}
\label{table:atomic-state}
\end{table}

\begin{enumerate}
    \item{The value of \ttt{index} must be unique over the molecular graph.}
    \item{If the value of \ttt{hydrogens} equals \ttt{Implicit}, then the values of \ttt{isotope}, \ttt{configuration}, and \ttt{charge} must equal their respective defaults. Furthermore, the value of \ttt{element} must equal one of \ttt{B}; \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; \ttt{S}; \ttt{F}; \ttt{Cl}; \ttt{Br}; \ttt{I}; or \ttt{None}.}
    \item{If the value of \ttt{selected} equals \ttt{true}, then the value of \ttt{element} must equal one of: \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; or \ttt{S}.}
\end{enumerate}

@@

\begin{table}
\caption{Bond State.}
\centering
\begin{tabular}{l L L l}
    \hline
    Attribute & Description & Type & Default \\
    \hline
    \ttt{index} & Unique identifier & \ttt{Index} & - \\
    \ttt{source} & Source atom index & \ttt{Index} & - \\
    \ttt{target} & Target atom index & \ttt{Index} & - \\
    \ttt{order} & Bond order & \ttt{Three} & 1 \\
    \ttt{elided} & Elision status & \ttt{boolean} & \ttt{false} \\
    \ttt{direction} & Partial conformational parity & \ttt{Option} \ttt{<Direction>} & \ttt{None} \\
    \hline
\end{tabular}
\label{table:bond-state}
\end{table}

\begin{enumerate}
    \item{The value of \ttt{index} must be unique over the molecular graph.}
    \item{The \ttt{source} and \ttt{target} attributes must reference atomic \ttt{index} attributes found elsewhere in the molecular graph.}
    \item{If the value of \ttt{direction} does not equal \ttt{None}, then the value of \ttt{order} must equal one.}
    \item{If the value of \ttt{elided} equals \ttt{true}, the values of \ttt{order} and \ttt{direction} must equal one and \ttt{None}, respectively.}
\end{enumerate}

@@

A molecular graph is described at the lowest level by its constitution. In Dialect, constitution comprises the molecular graph and those attributes locating nucleons and electrons at specific atoms and bonds. All other attributes lie outside the scope of constitution. Atomic attributes related to constitution are: \ttt{index}; \ttt{element}; \ttt{hydrogens}; and \ttt{charge}. Bond attributes related to constitution are \ttt{source}, \ttt{target}, \ttt{order}, and \ttt{elided}.

>>>>

\section*{Goals and Overall Design}

@@

Dialect molecules inherit behavior from an underlying \textit{molecular tree}. A tree is a data structure comprised of nodes. Each node is associated with zero or more child nodes through a directed edge. A molecular tree is a specialized tree onto which chemically-relevant metadata have been overlaid. The presence of this overlay allows nodes and edges to represent atoms and bonds, respectively.

A molecular tree is related to the more widely-used concept of a \textit{molecular graph}~\cite{balaban:1985}. There are two main difference: (1) bonds are undirected in molecular graphs but directed in molecular trees; and (2) cycles are possible in molecular graphs but not in molecular trees. The difference might appear to render trees unsuitable to represent molecules, in which cycles often play a prominent structural role. But as the next sections demonstrate, the differences between molecular trees and graphs are not so large as they might at first appear.

A Dialect molecule is a molecular tree having zero or more nodes and zero or more edges. The empty molecule, devoid of atoms and bonds, is therefore allowed. Dialect imposes no upper bound on the number of nodes or edges, although limitations in hardware and software may impose a practical upper bound on molecular size.

@@

Dialect uses several abstractions collectively known as the \textit{data model}. The data model defines metadata attributes associated with the components of a molecular tree. These attributes are themselves composed of \textit{primitives} (Table~\ref{table:primitives}), or irreducible data types.

% TODO: completely re-work this to fit with the exact model being used by the reference implementation

\begin{table}
\begin{minipage}{\textwidth}
\renewcommand*\footnoterule{}
\caption{Data Model Primitives.}
\label{table:primitives}
\centering
\begin{tabular}{l l l}
    \hline
    Name & Description & Values\footnote{Ranges defined as \ttt{a..b} include \ttt{a} and exclude \ttt{b}.} \\
    \hline
    \ttt{Option<T>} & Optional Value & \ttt{None}, \ttt{T} \\
    \ttt{Three} & Integer count & 1..4 \\
    \ttt{Dec} & Integer count &0..10 \\
    \ttt{SignedDec} & Integer count & -9..10 \\
    \ttt{Thou} & Integer count & 1..1000 \\
    \ttt{Element} & IUPAC-\-approved element symbol & \ttt{Ac}, \ttt{Ag}, \ttt{Al}, {\ldots}\ttt{Zn}\\
    \ttt{Configuration} & Configurational descriptor & \ttt{Right}, \ttt{Left} \\
    \ttt{Direction} & Conformational descriptor & \ttt{Up}, \ttt{Down} \\
    \ttt{Implicit} & Implicit hydrogen count flag & - \\
    \ttt{boolean} & Boolean & \ttt{true}, \ttt{false} \\
    \hline
\end{tabular}
\end{minipage}
\end{table}

@@

\begin{table}
\caption{Atomic State.}
\centering
\begin{tabular}{l L L l}
    \hline
    Attribute & Description & Type & Default \\
    \hline
    \ttt{element} & Elemental symbol & \ttt{Option} \ttt{<Element>} & \ttt{None} \\
    \ttt{hydrogens} & Hydrogen count & \ttt{Implicit}, \ttt{Dec} & 0 \\
    \ttt{configuration} & Configuration & \ttt{Option} \ttt{<Configuration>} & \ttt{None} \\
    \ttt{charge} & Formal charge & \ttt{SignedDec} & 0 \\
    \ttt{isotope} & Mass number & \ttt{Option<Thou>} & \ttt{None} \\
    \ttt{selected} & Selection status & \ttt{boolean} & \ttt{false} \\
    \hline
\end{tabular}
\label{table:atomic-state}
\end{table}

\begin{enumerate}
    \item{If the value of \ttt{hydrogens} equals \ttt{Implicit}, then the values of \ttt{isotope}, \ttt{configuration}, and \ttt{charge} must equal their respective defaults. Furthermore, the value of \ttt{element} must equal one of \ttt{B}; \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; \ttt{S}; \ttt{F}; \ttt{Cl}; \ttt{Br}; \ttt{I}; or \ttt{None}.}
    \item{If the value of \ttt{selected} equals \ttt{true}, then the value of \ttt{element} must equal one of: \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; or \ttt{S}.}
\end{enumerate}

@@

\begin{table}
\caption{Bond State.}
\centering
\begin{tabular}{l L L l}
    \hline
    Attribute & Description & Type & Default \\
    \hline
    \ttt{order} & Bond order & \ttt{Three} & 1 \\
    \ttt{elided} & Elision status & \ttt{boolean} & \ttt{false} \\
    \ttt{direction} & Partial conformational parity & \ttt{Option} \ttt{<Direction>} & \ttt{None} \\
    \ttt{target} & Child atom & \ttt{Atom} & - \\
    \hline
\end{tabular}
\label{table:bond-state}
\end{table}

\begin{enumerate}
    \item{If the value of \ttt{direction} does not equal \ttt{None}, then the value of \ttt{order} must equal one.}
    \item{If the value of \ttt{elided} equals \ttt{true}, the values of \ttt{order} and \ttt{direction} must equal one and \ttt{None}, respectively.}
\end{enumerate}

@@

A molecular tree is described at the lowest level by its constitution. In Dialect, constitution comprises those attributes locating nucleons and electrons at specific atoms and bonds. All other attributes lie outside the scope of constitution. Atomic attributes related to constitution are: \ttt{element}; \ttt{hydrogens}; and \ttt{charge}. Bond attributes related to constitution are \ttt{order} and \ttt{elided}.
